Data Specialization for a Staged Lambda Calculus

### 2015-02-20 paper/splitting.tex
mention fixpoints; change first paragraph of section 4

The primary strength of \lang\ is the expressive power afforded by
its ability to syntactically interleave first- and second-stage computations,
as well as its ability to form mixed-stage abstractions.

In this section, we provide a {\em splitting algorithm} to translate the multistage terms of \lang\
into a simpler but equivalent form.
The output of splitting is simpler because it is {\em syntactically separated}; that is, 
the first and second stages are represented by distinct subterms and are only connected at the top syntactic level.
The output of splitting is equivalent (to its input) because it can be evaluated to the same residual 
that would be produced by the semantics of \ref{sec:semantics}.

More precisely, for any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e q$),
splitting translates $e$ into some syntactically separated program $\mathcal{P}$ 
(via the new relation $e \splittwosym \mathcal{P}$), which
can also be reduced to $q$ (via the new relation $\mathcal{P} \sepredtwosym q$).
We enforce the ``syntactically separated'' condition by saying that $\mathcal{P}$ 
can only have one form, namely $\pipeS p l r$, where  
$p$ (called the {\em precomputation}) is a monostage term encoding all the
first-stage computation that was in $e$, 
and $l.r$ (called the {\em resumer}) is a monostage term encoding all the
second-stage computation that was in $e$.  
Likewise, $\sepredtwosym$ is defined by a single rule, which evaluates the precomputation and plugs it in for $l$:
\[
\infer{\sepredtwo {\pipeS p l r} {[b/l]r}} {\reduce p b}
\]
Here we call $b$ the {\em boundary value}, as it represents the communication at the boundary between stages.
A summary of these operations and the relationship between them is provided on the right side of \ref{fig:splittingSummary}.

>>>>

Although the type system of \lang\ ensures that we can perform all first-stage
computations before any second-stage ones, actually evaluating terms in this
fashion is somewhat intricate, because stages are syntactically interleaved
within terms. In this section, we present a \emph{splitting algorithm} which
statically converts a term in \lang\ into a pair of 
syntactically % I don't like this word but temporarily here for continuity
separate programs corresponding to the two stages of computation in that term.

%The output of splitting is simpler because it is {\em syntactically separated};
%The output of splitting is equivalent (to its input) because it can be
%evaluated to the same residual that would be produced by the semantics of
%\ref{sec:semantics}.

More precisely, for any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e q$),
$e$ splits into a syntactically separated program $\mathcal{P}$ 
(via the new relation $e \splittwosym \mathcal{P}$), which
can also be reduced to $q$ (via the new relation $\mathcal{P} \sepredtwosym q$).
We enforce the ``syntactically separated'' condition by saying that $\mathcal{P}$ 
can only have one form, namely $\pipeS p l r$, where  
$p$ (called the {\em precomputation}) is a monostage term encoding all the
first-stage computation that was in $e$, 
and $l.r$ (called the {\em resumer}) is a monostage term encoding all the
second-stage computation that was in $e$.  
Likewise, $\sepredtwosym$ is defined by a single rule, which evaluates the precomputation and plugs it in for $l$:
\[
\infer{\sepredtwo {\pipeS p l r} {[b/l]r}} {\reduce p b}
\]
Here we call $b$ the {\em boundary value}, as it represents the communication at the boundary between stages.
A summary of these operations and the relationship between them is provided on the right side of \ref{fig:splittingSummary}.
